Meta-Analysis of the `Ironic' Effects of Intergroup Contact

### 2021-04-16 manuscript/manuscript.tex
add abstract

\maketitle

@@

As a growing number of studies had been examining these relationships,
we conducted a preregistered meta-analytic review of the available
evidence for the hypothesized negative effects of intergroup contact on
perceived injustice, collective action, and policy support in
disadvantaged groups.

Our first objective was to systematically review the available evidence
and to evaluate its strengths and limitations. Our second objective was
to synthesize the available evidence to establish the direction and
magnitude of the relationships between intergroup contact and the three
outcomes. Doing so allows us, first, to determine whether the available
evidence supports a negative or positive (see Pettigrew et al., 2011)
relationship between contact and support for social change and, second,
to estimate the strength of this relationship. Our third objective was
to explore the variability of this relationship across study settings,
designs, and other potential moderators. For example, we examined
whether any effects were confined to Western, Educated, Industrialized,
Rich, Democratic (WEIRD) settings, the cultural context of most
psychological research (Henrich et al., 2010). Our fourth objective was
to assess how robust the available evidence was to publication bias.
Together, these objectives provided a comprehensive answer to the
question whether intergroup contact diminishes support for social change
in disadvantaged groups.

In addition, we explored the evidence for two alternative accounts of
this relationship. First, we considered studies that measured both
ingroup and outgroup contact as predictors of the three outcomes.
Sengupta et al. (2015) argued that contact with other members of the
disadvantaged ingroup would increase political mobilization in
disadvantaged groups. As outgroup contact places limits on how much
ingroup contact disadvantaged-group members can have (Pfister et al.,
2020), the mobilizing effect of ingroup contact is an alternative
explanation for negative associations between outgroup contact and
support for social change.

Second, we considered studies that measured both negative and positive
(outgroup) contact as predictors of the three outcomes. Researchers
(Hayward et al., 2018; Reimer et al., 2017) found negative contact with
advantaged-group members to be associated with greater perceived
discrimination and stronger collective action intentions. Reimer et al.
(2017) argued that, to the extent disadvantaged-group members who have
more positive contact also tend to have less negative contact, the
mobilizing effect of negative contact is an alternative explanation for
negative associations between outgroup contact and support for social
change. For both alternative explanations, we collate and synthesize the
available evidence for the relative effects of each predictor variable
on the three outcomes.

@@

We prepared a protocol in accordance with the PRISMA-P statement (Moher
et al., 2015) and preregistered it on the Open Science
Framework.\footnote{\url{https://osf.io/ryuev/?view_only=8938ac6178914da98bdda1da4a6e27f0}}
Preregistration precludes undisclosed flexibility in study selection,
outcome selection, and other decisions that influence effect size
estimates---and thus prevents confirmation bias from determining the
conclusions of a meta-analysis (Lakens et al., 2016). We report all
deviations from our protocol in the Supplemental Materials. We make all
data and analysis scripts available online.\footnote{\url{https://osf.io/w5tqv/?view_only=5cbf7ad5b2b449a7a0f4ada0337b802a}}

@@

Figure 1 shows a flow diagram illustrating our search strategy, study
selection, and data collection. Our preregistered search strategy
returned 2,610 unique records from electronic databases. Of these, we
excluded 2,379 (91\%) ineligible records after screening titles,
abstracts, and keywords. Of the remaining 231 records, we excluded 116
(50\%) ineligible records after reviewing full-text manuscripts. We
supplemented these records with 14 unpublished studies solicited from
researchers and 6 records that cited at least three relevant works. Of
135 eligible studies, we had to exclude 39 (29\%) studies for which we
could not extract or impute any relevant effect size. At each stage, we
also excluded studies that used the same data as another study. Our
final sample comprised effect sizes from \(J = 96\) studies spanning
\(N = 211,360\) participants in \(I = 137\) samples.

Figure 2 provides a qualitative overview of the literature. Even though
most participants came from studies conducted in India (\(N = 104,639\))
and the United States (\(N = 61,089\); Figure 2a), most samples were
collected in North America (\(I = 48\)) and Europe (\(I = 40\)), with
few samples from South America and Africa (Figure 2b). Most samples
focused on relative inequalities that resulted from long-term migration
(\(I = 54\)), slavery (\(I = 19\)), colonization (\(I = 17\)), and
short-term migration (\(I = 14\)). With few exceptions, studies used
observational, cross-sectional survey designs. Samples sizes ranged from
64 to 49,764 (\(\textit{Mdn} = 282\)). Unlike most psychological
research, 2 in 5 samples used probability or representative sampling. Of
all samples, 55\% (\(I = 75\)) were collected with the intention to
examine the effects of intergroup contact on the outcomes considered in
this meta-analysis.\footnote{Many studies examined acculturation
  processes (Berry, 1997) and measured both intergroup contact and
  perceived injustice without the relationship between the two being of
  interest.} Together, these observations highlight both limitations and
strengths of the empirical literature.

@@

Figure 7 shows adjusted and unadjusted estimates with confidence
intervals for the three outcomes. Comparing adjusted and unadjusted
estimates shows that the PET-PEESE and 3PSM methods tended to estimate
the mean correlation coefficients to be closer to zero than the
unadjusted estimate, with all confidence intervals including zero. The
\emph{p}-uniform* method and the subgroup analysis tended to estimate
the mean correlations coefficients to be closer to the unadjusted
estimates, with the confidence interval for perceived injustice---but
not for the other outcomes---excluding zero. All methods, however,
resulted in confidence intervals that largely overlapped with, but were
wider than, the confidence intervals around the unadjusted estimates.
This reflects the reduced power of the various methods to correct for
meta-biases when sample sizes are small, effect sizes are heterogeneous,
or publication bias is strong (Carter et al., 2019). Therefore, there
was insufficient evidence to conclude \emph{either} that, after
adjusting for publication bias, intergroup contact was not associated
with support of social change in disadvantaged groups \emph{or} that
publication bias did not affect the relevant literature.

@@

There is an emerging consensus that intergroup contact has the `ironic'
effect of undermining support for social change in disadvantaged groups.
We conducted a preregistered meta-analytic test of this effect across 96
studies with 137 samples of 211,360 disadvantaged-group members. We
found that, based on the available evidence, the associations of
intergroup contact with perceived injustice (\(\Pr (r < 0) > 99.9\%\)),
collective action (\(\Pr (r < 0) = 93.6\%\)), and support for reparative
policies (\(\Pr (r < 0) = 98.1\%\)) were, on average, much more likely
to be negative than positive. Thus, our meta-analysis seems to support
the emerging consensus that intergroup contact diminishes support for
social change in disadvantaged groups. We use the rest of this section
to argue why this conclusion might be premature.

First, the estimated effect sizes for the average associations of
intergroup contact with perceived injustice (\(r = -.07\)), collective
action (\(r = -.06\)), and policy support (\(r = -.07\)) were small.
Across the three outcomes, the 95\% most plausible estimates included
effect sizes between \(r = -.14\) and \(r = .02\). Effect sizes were
thus much smaller than for the association between contact and prejudice
in minority (\(r = -.18\)) and majority (\(r = -.23\)) groups (Tropp \&
Pettigrew, 2005).\footnote{Comparing standardized effect sizes assumes
  that the two outcomes are of equal importance. That is not always the
  case: for example, a drug that reduces mild symptoms by 0.20 standard
  deviations is not better than a drug that reduces deaths by 0.10
  standard deviations.} Small effects can still be important if they
accumulate over time or across people. For example, even a small change
in policy attitudes could, if it affects enough people, sway a tight
election. Cumulative effects, however, should not be assumed without an
empirical or theoretical rationale (Funder \& Ozer, 2019)---neither of
which has been provided in the published literature.

@@

\begin{figure}
\centering
\caption{Directed acyclic graphs for plausible causal relations between intergroup contact and support for social change}
\includegraphics[scale=1]{../figures/figure-9}
\caption*{\textit{Note.} IC = (Positive) Intergroup Contact, SC = Support for Social Change, U = Unobserved Variable(s), NC = Negative (Integroup) Contact, IN = Ingroup Contact}
\label{fig:f9}
\end{figure}

Third, the available evidence is almost entirely from cross-sectional,
observational studies and thus consistent with alternative explanations
for the observed associations (Figure 9). Support for social change
could reduce intergroup contact (B) rather than the other way around
(A). For example, disadvantaged-group members involved in collective
action might purposefully avoid forming friendships with
advantaged-group members. Alternatively, the association between
intergroup contact and support for social change could be spurious with
both being caused by an unobserved confounder (C). For example, a
disadvantaged-group member's socioeconomic status might both expose them
to more advantaged-group members and reduce their perception of
injustice. Future research should prioritize longitudinal studies to
confirm the direction of the observed associations and (field)
experiments to rule out confounding and other alternative
explanations.\footnote{Published research includes only three
  longitudinal studies (Koschate et al., 2012; Reimer et al., 2017;
  Tropp et al., 2012), all of which conflated within-person change and
  between-person stability (Hamaker et al., 2015), two experimental
  studies with a no-contact control condition (Becker et al., 2013;
  Droogendyk et al., 2016), and two intervention studies (Reimer et al.,
  2021; Shani \& Boehnke, 2017).}

Further, the observed association between intergroup contact and support
for social change could be confounded by an alternative predictor that
\emph{increases} support for social change (D, E). Supporting Reimer et
al.'s (2017) argument, we found that positive intergroup contact was not
associated with support for social change after controlling for negative
contact which, in turn, was associated with \emph{greater} support for
social change. We found mixed evidence for ingroup contact as an
alternative explanation for the `ironic' effects of intergroup contact
(Sengupta et al., 2015) but note that few studies measured both ingroup
and outgroup contact. Going forward, researchers should clarify which
aspects of intergroup contact they hypothesize to affect support for
social change---and include measures that allow testing both the
hypothesized and competing explanations.

Fourth, we cannot rule out that biases in the literature caused our
meta-analysis to overestimate the `ironic' effects of intergroup
contact. Publication bias could have prevented studies that found
positive or non-significant associations from entering the published
literature. Even though we used state-of-the-art methods to detect and
correct for publication bias, we did not find conclusive evidence for or
against publication bias. That said, 61\% of samples were \emph{either}
not collected with the intention to study the effects of intergroup
contact on a relevant outcome (45\%) \emph{or} from unpublished studies
(32\%)---and thus unlikely to have been affected by publication bias.
Another bias that could have affected our findings is that researchers
made data-dependent decisions when analyzing data that inflate
false-positive findings (Gelman \& Loken, 2014). Future research should
use preregistration and other methods to prevent undisclosed flexibility
in data collection and analysis.\footnote{To date, one preregistered
  study has been published (Hässler et al., 2020).}

@@

>>>>

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
    \noindent Critics have argued that intergroup contact, psychology's
most-researched paradigm for reducing prejudice, has the `ironic' effect
of stifling support for social change in disadvantaged groups. We
conducted a preregistered meta-analytic test of this effect across 96
studies with 137 samples of 211,360 disadvantaged-group members. As
predicted, intergroup contact was, on average, associated with less
perceived injustice (\(r = -.07\)), collective action (\(r = -.06\)),
and support for reparative policies (\(r = -.07\)) in disadvantaged
groups. However, these associations were small, variable, and consistent
with alterative explanations. Moderator analyses found that these
associations were strongest in studies on short-term migration and
post-colonial relations, that measured intergroup contact directly, and
that involved adults. After controlling for the positive association of
negative contact with support for social change, positive contact was
not (negatively) associated with any outcome in studies that measured
both predictors. We close by highlighting open questions about the
relation between intergroup contact and social change.
    \end{abstract}
  \end{@twocolumnfalse}
]

@@

We conducted a preregistered meta-analysis of the evidence for the
hypothesized negative effects of intergroup contact on perceived
injustice, collective action, and policy support in disadvantaged
groups. Our first objective was to systematically review the available
evidence and to evaluate its strengths and limitations. Our second
objective was to synthesize the available evidence to establish the
direction and magnitude of the relationships between intergroup contact
and the three outcomes. Doing so allowed us, first, to determine whether
the available evidence supports a negative or positive (see Pettigrew et
al., 2011) relationship between contact and support for social change
and, second, to estimate the strength of this relationship. Our third
objective was to explore the variability of this relationship across
study settings, designs, and other potential moderators. For example, we
examined whether any effects were confined to Western, Educated,
Industrialized, Rich, Democratic (WEIRD) settings, the cultural context
of most psychological research (Henrich et al., 2010). Our fourth
objective was to assess how robust the available evidence was to
publication bias.

In addition, we collated the evidence for two alternative accounts of
this relationship. First, we considered studies that measured both
ingroup and outgroup contact as predictors of the three outcomes.
Sengupta et al. (2015) argued that contact with other members of the
disadvantaged ingroup would increase political mobilization in
disadvantaged groups. As outgroup contact places limits on how much
ingroup contact disadvantaged-group members can have (Pfister et al.,
2020), the mobilizing effect of ingroup contact is an alternative
explanation for negative associations between outgroup contact and
support for social change. Second, we considered studies that measured
both negative and positive (outgroup) contact as predictors of the three
outcomes. Researchers (Hayward et al., 2018; Reimer et al., 2017) found
negative contact with advantaged-group members to be associated with
greater perceived discrimination and stronger collective action
intentions. Reimer et al. (2017) argued that, to the extent
disadvantaged-group members who have more positive contact also tend to
have less negative contact, the mobilizing effect of negative contact is
an alternative explanation for negative associations between outgroup
contact and support for social change.

@@

We prepared a protocol in accordance with the PRISMA-P statement (Moher
et al., 2015) and preregistered it on the Open Science
Framework.\footnote{\url{https://osf.io/ryuev/?view_only=8938ac6178914da98bdda1da4a6e27f0}}
Preregistration precludes undisclosed flexibility in study selection,
outcome selection, and other decisions that influence effect size
estimates---and thus prevents confirmation bias from determining the
conclusions of a meta-analysis (Lakens et al., 2016). We report
deviations from our protocol in the Supplemental Materials. We make data
and analysis scripts available online.\footnote{\url{https://osf.io/w5tqv/?view_only=5cbf7ad5b2b449a7a0f4ada0337b802a}}

@@

Figure 1 shows a flow diagram illustrating our search strategy, study
selection, and data collection. Our preregistered search strategy
returned 2,610 unique records from electronic databases. Of these, we
excluded 2,379 (91\%) ineligible records after screening titles,
abstracts, and keywords. Of the remaining 231 records, we excluded 116
(50\%) ineligible records after reviewing full-text manuscripts. We
supplemented these records with 14 unpublished studies solicited from
researchers and 6 records that cited at least three relevant works. Of
135 eligible studies, we had to exclude 38 (28\%) studies for which we
could not extract or impute any relevant effect size. At each stage, we
also excluded studies that used the same data as another study. Our
final sample comprised effect sizes from \(J = 96\) studies spanning
\(N = 211,360\) participants in \(I = 137\) samples.

Figure 2 provides a qualitative overview of the literature. Even though
most participants came from studies conducted in India (\(N = 104,639\))
and the United States (\(N = 60,983\); Figure 2a), most samples were
collected in North America (\(I = 48\)) and Europe (\(I = 40\)), with
few samples from South America and Africa (Figure 2b). Most samples
focused on relative inequalities that resulted from long-term migration
(\(I = 56\)), slavery (\(I = 19\)), colonization (\(I = 19\)), and
short-term migration (\(I = 14\)). With few exceptions, studies used
observational, cross-sectional survey designs. Samples sizes ranged from
64 to 49,764 (\(\textit{Mdn} = 282\)). Unlike most psychological
research, 2 in 5 samples used probability or representative sampling. Of
all samples, 55\% (\(I = 75\)) were collected with the intention to
examine the effects of intergroup contact on the outcomes considered in
this meta-analysis.\footnote{Many studies examined acculturation
  processes (Berry, 1997) and measured both intergroup contact and
  perceived injustice without the relationship between the two being of
  interest.} Together, these observations highlight both limitations and
strengths of the empirical literature.

@@

Figure 7 shows adjusted and unadjusted estimates with confidence
intervals for the three outcomes. Comparing adjusted and unadjusted
estimates shows that the PET-PEESE and 3PSM methods tended to estimate
the mean correlation coefficients to be closer to zero than the
unadjusted estimate, with all confidence intervals including zero. The
\emph{p}-uniform* method and the subgroup analysis tended to estimate
the mean correlations coefficients to be closer to the unadjusted
estimates, with the confidence interval for perceived injustice---but
not for the other outcomes---excluding zero. All methods, however,
resulted in confidence intervals that largely overlapped with, but were
wider than, the confidence intervals around the unadjusted estimates.
This reflects the reduced power of the various methods to correct for
meta-biases when sample sizes are small, effect sizes are heterogeneous,
or publication bias is strong (Carter et al., 2019). Therefore, we did
not find evidence for publication bias---but also did not find
conclusive evidence against it.

@@

There is an emerging consensus that intergroup contact has the `ironic'
effect of undermining support for social change in disadvantaged groups.
We conducted a preregistered meta-analytic test of this effect across 96
studies with 137 samples of 211,360 disadvantaged-group members. We
found that, based on the available evidence, the associations of
intergroup contact with perceived injustice (\(\Pr (r < 0) > 99.9\%\)),
collective action (\(\Pr (r < 0) = 93.6\%\)), and support for reparative
policies (\(\Pr (r < 0) = 98.1\%\)) were, on average, much more likely
to be negative than positive. Thus, our meta-analysis seems to support
the emerging consensus. We use the rest of this section to argue why
this conclusion might be premature.

First, the estimated effect sizes for the average associations of
intergroup contact with perceived injustice (\(r = -.07\)), collective
action (\(r = -.06\)), and policy support (\(r = -.07\)) were small.
Across the three outcomes, the 95\% most plausible estimates included
effect sizes between \(r = -.14\) and \(r = .02\). Effect sizes were
thus much smaller than for the association between contact and prejudice
in minority (\(r = -.18\)) and majority (\(r = -.23\)) groups (Tropp \&
Pettigrew, 2005).\footnote{Comparing standardized effect sizes assumes
  that the two outcomes are of equal importance. That is not always the
  case: for example, a drug that reduces mild symptoms by 0.20 standard
  deviations is not better than a drug that reduces deaths by 0.10
  standard deviations.} Small effects can still be important if they
accumulate over time or across people. For example, even a small change
in policy attitudes could, if it affects enough people, sway a tight
election. Cumulative effects, however, should not be assumed without an
empirical or theoretical rationale (Funder \& Ozer, 2019).

@@

Third, the available evidence is almost entirely from cross-sectional,
observational studies and thus consistent with alternative explanations
for the observed associations. Support for social change could reduce
intergroup contact rather than the other way around. For example,
disadvantaged-group members involved in collective action might
purposefully avoid forming friendships with advantaged-group members.
Alternatively, the association between intergroup contact and support
for social change could be spurious with both being caused by an
unobserved confounder. For example, a disadvantaged-group member's
socioeconomic status might both expose them to more advantaged-group
members and reduce their perception of injustice. Future research should
prioritize longitudinal studies to confirm the direction of the observed
associations and (field) experiments to rule out confounding and other
alternative explanations.\footnote{Published research includes only
  three longitudinal studies (Koschate et al., 2012; Reimer et al.,
  2017; Tropp et al., 2012), all of which conflated within-person change
  and between-person stability (Hamaker et al., 2015), two experimental
  studies with a no-contact control condition (Becker et al., 2013;
  Droogendyk et al., 2016), and two intervention studies (Reimer et al.,
  2021; Shani \& Boehnke, 2017).}

Further, the observed association between intergroup contact and support
for social change could be confounded by an alternative predictor that
\emph{increases} support for social change. Supporting Reimer et al.'s
(2017) argument, we found that positive intergroup contact was not
associated with support for social change after controlling for negative
contact which, in turn, was associated with \emph{greater} support for
social change. We found mixed evidence for ingroup contact as an
alternative explanation for the `ironic' effects of intergroup contact
(Sengupta et al., 2015) but note that few studies measured both ingroup
and outgroup contact. Going forward, researchers should clarify which
aspects of intergroup contact should theoretically affect support for
social change---and include measures that allow testing both the
hypothesized and competing explanations.

Fourth, we cannot rule out that biases in the literature caused our
meta-analysis to overestimate the `ironic' effects of intergroup
contact. Publication bias could have prevented studies that found
positive or non-significant associations from entering the published
literature. We used state-of-the-art methods to detect and correct for
publication bias, yet did not find conclusive evidence for or against
publication bias. That said, 61\% of samples were either not collected
with the intention to study the effects of intergroup contact on a
relevant outcome (45\%) or from unpublished studies (32\%)---and thus
unlikely to have been affected by publication bias. Another bias that
could have affected our findings is that researchers made data-dependent
decisions when analyzing data that inflate false-positive findings
(Gelman \& Loken, 2014). Future research should use preregistration and
other methods to prevent undisclosed flexibility in data collection and
analysis.\footnote{To date, one preregistered study has been published
  (Hässler et al., 2020).}

@@

\leavevmode\hypertarget{ref-berry_immigration_1997}{}%
Berry, J. W. (1997). Immigration, acculturation, and adaptation.
\emph{Applied Psychology}, \emph{46}(1), 5--34.
\url{https://doi.org/10.1111/j.1464-0597.1997.tb01087.x}
